---
title: "MStat Project"
author: "Annie Fuller"
date: "3/10/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('spatstat')
require('RCurl')
```

$\hat{\Sigma}_{GR} = |B(1)|\frac{r^{q+2}}{q+2}I_q - \frac{1}{|B(R-r)|}\sum_{i\in N_0}\sum_{j \in N_i}(X_i - X_j)(X_i-X_j)'$

Below is the original R code sourced from in Gaussian determinantal processes: A new model for directionality in data by Subhroshekhar Ghosh and Philippe Rigollet

Downloaded on March 15, 2022 at https://klein.mit.edu/~rigollet/Code/DPP/DPPcode.R

```{r, Sigma_GR function}
norm_vec <- function(x){
  sqrt(sum(x^2))
}

sigmat <- function(X, r, att=1,R){
  #Inputs are X: dataframe where each row is a point in d dimensions.
  #r>0 is the scale at which the problem is looked at default is sqrt(d), and R is the radius of the window in which the points are.
  # if no r, it is attenuated by the factor att
  n<-dim(X)[1]
  d<-dim(X)[2]
  if(missing(r)){r <- sqrt(d)}
  r <- r*att
  print(r)
  sigma=matrix(0, nrow=d,ncol=d)
  vol <- abs(((pi^(d/2))/(gamma(d/2 + 1)))*(R-r)^d)
  print(vol)

  vec_dimn <- integer(n)
    
for (i in 1:n){
  dist <- sqrt(rowSums(mapply(function(x,y) (x-y)^2, X, X[i,])))
  index_neighbors <- (dist<r)&(dist>0)
  neighbors <- as.matrix(X[index_neighbors,])
  nn <- dim(neighbors)[1]
  vec_dimn[i] <- nn
  ones <- cbind(rep(1,nn));
  xi <- rbind(as.numeric(X[i,]))
  neighbors <- neighbors - ones%*%xi
  sigma=1/vol*(sigma+t(neighbors)%*%neighbors) #divided by n added to appropriately scale eigenvalues
  
}
  s <- svd(sigma)
  ss <- s$v
  S <- ss[,1:2] #project onto the space spanned by first two eigenvectors of estimator
  Y=list("pro"=as.matrix(X)%*%S, "s"=s,"sigma"=as.matrix(sigma), "dimn"=vec_dimn,"max_dist"=max(dist))
  
  return(Y)
}

sigmat_no_sc <- function(X, r, att=1){
  #Inputs are X: dataframe where each row is a point in d dimensions.
  #r>0 is the scale at which the problem is looked at default is sqrt(d), and R is the radius of the window in which the points are.
  # if no r, it is attenuated by the factor att
  n<-dim(X)[1]
  d<-dim(X)[2]
  if(missing(r)){r <- sqrt(d)}
  r <- r*att
  sigma=matrix(0, nrow=d,ncol=d)

  vec_dimn <- integer(n)
    
for (i in 1:n){
  dist <- sqrt(rowSums(mapply(function(x,y) (x-y)^2, X, X[i,])))
  index_neighbors <- (dist<r)&(dist>0)
  neighbors <- as.matrix(X[index_neighbors,])
  nn <- dim(neighbors)[1]
  vec_dimn[i] <- nn
  ones <- cbind(rep(1,nn));
  xi <- rbind(as.numeric(X[i,]))
  neighbors <- neighbors - ones%*%xi
  sigma=(sigma+t(neighbors)%*%neighbors) #divided by n added to appropriately scale eigenvalues
  
}
  s <- svd(sigma)
  ss <- s$v
  S <- ss[,1:2] #project onto the space spanned by first two eigenvectors of estimator
  Y=list("pro"=as.matrix(X)%*%S, "s"=s,"sigma"=as.matrix(sigma), "dimn"=vec_dimn,"max_dist"=max(dist))
  
  return(Y)
}

```

```{r,import data}
#import data
```{r, import data}
library('tidyr')

#iris data
iris_clus <- read.csv("iris.data",header=F)
iris <- as.data.frame(iris[,c(1:4)])
iris_scaled <- as.data.frame(scale(iris,scale=TRUE,center=TRUE))

#WDBC data
#https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data

wdbc <- read.csv("wdbc.data", sep=",", header = F)

wdbc[wdbc == "?"] <- NA
#which(is.na(wdbc))
wdbc_clus <- as.data.frame(wdbc)
wdbc <- as.data.frame(wdbc[,c(3:32)])
# wdbc <- as.data.frame(scale(wdbc,scale=FALSE,center=FALSE)) #unscaled data
wdbc_scaled <- as.data.frame(scale(wdbc,scale=TRUE,center=TRUE)) #scale data

#ecoli
ecoli <- read.csv("ecoli.data",sep="",header=F)
ecoli <- subset(ecoli, V9 != "omL" & V9 != "imL" & V9 != "imS") 
#removing categories with only a few observations
ecoli <- ecoli[,c(1:3,6:9)]
#removing V4 and V5 as they only have one or two observation options
ecoli_clus <- ecoli
ecoli <- ecoli[,c(2:6)]
ecoli_scaled <- as.data.frame(scale(ecoli,scale=TRUE,center=TRUE))
```

```{r,r experiments}
r_experiment <- .75

#wdbc: pca vs. Sigma_GR eigenvector projections
pca_wdbc_scaled <- prcomp(wdbc_scaled, scale=TRUE, center=TRUE)
plot((pca_wdbc_scaled$sdev)^2,main="Standard PCA",ylab="Eigenvalues",xlab="Index")
# screeplot(pca_wdbc,npcs=length(pca_wdbc$sdev), type="lines")

#recreating article
wdbc_GR_article <- sigmat_no_sc(wdbc_scaled,r=24.76723,)
plot(wdbc_GR_article$s$d,main="DPP Method, r=10",ylab="Eigenvalues",xlab="Index")

#using r=.75
wdbc_GR_75 <- sigmat_no_sc(wdbc_scaled,r=r_experiment)
plot(wdbc_GR_75$s$d,main="DPP Method, r=0.75",ylab="Eigenvalues",xlab="Index")

#default r (sqrt of q)
wdbc_GR_default <- sigmat_no_sc(wdbc_scaled)
plot(wdbc_GR_default$s$d,main="DPP Method, default r",ylab="Eigenvalues",xlab="Index")
```

```{r, determine R}
#determine R
determine_R_2 <- function(X) {
  cent <- mapply(mean, X)
  dist1 <- sqrt(rowSums(mapply(function(x,y) (x-y)^2, X, cent)))
  return(max(dist1))
}

iris_R <- determine_R_2(iris_scaled)
wdbc_R <- determine_R_2(wdbc_scaled)
ecoli_R <- determine_R_2(ecoli_scaled)
```

```{r, PCA vs. DPP & screeplots}
#iris: pca vs. Sigma_GR eigenvalues
pca_iris <- prcomp(iris_scaled, scale=TRUE, center=TRUE)
jpeg("pca_iris.jpg")
plot((pca_iris$sdev)^2,main="Iris: Standard PCA",ylab="Eigenvalues",xlab="Index",type="b")
dev.off()

#DPP method
iris_GR <- sigmat(iris_scaled,R=iris_R)
jpeg("DPP_iris.jpg")
plot(iris_GR$s$d,main="Iris: DPP Method, default r",ylab="Eigenvalues",xlab="Index",type="b")
dev.off()

#--
  
#wdbc: pca vs. Sigma_GR eigenvector projections
pca_wdbc <- prcomp(wdbc_scaled, scale=TRUE, center=TRUE)
jpeg("pca_wdbc.jpg")
plot((pca_wdbc$sdev)^2,main="WDBC: Standard PCA",ylab="Eigenvalues",xlab="Index",type="b")
dev.off()

#DPP method
wdbc_GR <- sigmat(wdbc_scaled,R=wdbc_R)
jpeg("DPP_wdbc.jpg")
plot(wdbc_GR$s$d,main="WDBC: DPP Method, default r",ylab="Eigenvalues",xlab="Index",type="b")
dev.off()

#--

#ecoli: pca vs. Sigma_GR eigenvector projections
jpeg("pca_ecoli.jpg")
pca_ecoli <- prcomp(ecoli_scaled,scale=TRUE,center=TRUE)
plot((pca_ecoli$sdev)^2,main="E.Coli: Standard PCA",ylab="Eigenvalues",xlab="Index",type="b")
dev.off()

#DPP method
ecoli_GR <- sigmat(ecoli_scaled,R=ecoli_R)
jpeg("DPP_ecoli.jpg")
plot(ecoli_GR$s$d,main="E.Coli: DPP Method, default r",ylab="Eigenvalues",xlab="Index",type="b")
dev.off()

```

```{r, projections}
library("factoextra")

#projection of first two principal components 
#vs. first two eigenvectors of Sigma_GR

#iris
jpeg("iris_pca_proj.jpg")
fviz_pca_ind(pca_iris, label="none", habillage=iris_clus$V5,title="Iris: Projection of first two PCs (standard PCA)",xlab="PCA-PC1",ylab="PCA-PC2")
#dev.off()

# iris_projection <- as.data.frame(iris_GR$pro)

b <- 250*(pi/180)
iris_rotation <- t(matrix(c(cos(b),-sin(b),sin(b),cos(b)),nrow=2))

iris_projection <- as.matrix(iris_GR$pro)
iris_projection <- t(iris_rotation%*%t(iris_projection))
iris_projection <- as.data.frame(iris_projection)

jpeg("iris_dpp_proj.jpg")
ggplot(iris_projection,aes(x=iris_projection[,1],y=iris_projection[,2],group=iris_clus$V5))+
  geom_point(aes(shape=iris_clus$V5,color=iris_clus$V5))+
  geom_vline(xintercept=0, linetype="longdash")+
  geom_hline(yintercept=0,linetype="longdash")+
  ggtitle("Iris: Projection of first two eigenvectors (DPP)") + 
  xlab("DPP-PC1") + ylab("DPP-PC2") + labs(color="Groups",shape="Groups") + 
  theme_bw() + theme(panel.border = element_blank())
dev.off()

#wdbc
jpeg("wdbc_pca_proj.jpg")
fviz_pca_ind(pca_wdbc, label="none", habillage=wdbc_clus$V2,title="WDBC: Projection of first two PCs (standard PCA)",xlab="PCA-PC1",ylab="PCA-PC2")
dev.off()

# wdbc_projection <- as.data.frame(wdbc_GR$pro)

a <- 45*(pi/180)
wdbc_rotation <- t(matrix(c(cos(a),-sin(a),sin(a),cos(a)),nrow=2))

wdbc_projection <- as.matrix(wdbc_GR$pro)
wdbc_projection <- t(wdbc_rotation%*%t(wdbc_projection))
wdbc_projection <- as.data.frame(wdbc_projection)

jpeg("wdbc_dpp_proj.jpg")
ggplot(wdbc_projection,aes(x=wdbc_projection[,1],y=wdbc_projection[,2],group=wdbc_clus$V2))+
  geom_point(aes(shape=wdbc_clus$V2,color=wdbc_clus$V2))+
  geom_vline(xintercept=0, linetype="longdash")+
  geom_hline(yintercept=0,linetype="longdash")+
  ggtitle("WDBC: Projection of first two eigenvectors (DPP)") + 
  xlab("DPP-PC1") + ylab("DPP-PC2") + labs(color="Groups",shape="Groups") + 
  theme_bw() + theme(panel.border = element_blank())
dev.off()

#ecoli
jpeg("ecoli_pca_proj.jpg")
fviz_pca_ind(pca_ecoli, label="none", habillage=ecoli_clus$V9,title="E.Coli: Projection of first two PCs (standard PCA)",xlab="PCA-PC1",ylab="PCA-PC2")
dev.off()

#ecoli_projection <- as.data.frame(ecoli_GR$pro)

c <- 1*(pi/180)
ecoli_rotation <- t(matrix(c(cos(c),-sin(c),sin(c),cos(c)),nrow=2))

ecoli_projection <- as.matrix(ecoli_GR$pro)
ecoli_projection  <- t(ecoli_rotation%*%t(ecoli_projection))
ecoli_projection <- as.data.frame(ecoli_projection)

jpeg("ecoli_dpp_proj.jpg")
ggplot(ecoli_projection,aes(x=-ecoli_projection[,1],y=ecoli_projection[,2],group=ecoli_clus$V9))+
  geom_point(aes(shape=ecoli_clus$V9,color=ecoli_clus$V9))+
  geom_vline(xintercept=0, linetype="longdash")+
  geom_hline(yintercept=0,linetype="longdash")+
  ggtitle("E.Coli: Projection of first two eigenvectors (DPP)") + 
  xlab("DPP-PC1") + ylab("DPP-PC2") + labs(color="Groups",shape="Groups") + 
  theme_bw() + theme(panel.border = element_blank())
dev.off()
```

```{r, projection test}
dpp_wdbc_test <- sigmat_no_sc(wdbc)

d <- 95*(pi/180)
test_rotation <- t(matrix(c(cos(d),-sin(d),sin(d),cos(d)),nrow=2))

test_projection <- as.matrix(dpp_wdbc_test$pro)
test_projection <- t(test_rotation%*%t(test_projection))
test_projection <- as.data.frame(test_projection)

jpeg("test_projection_wdbc.jpg")
ggplot(test_projection,aes(x=-test_projection[,1],y=test_projection[,2],group=wdbc_clus$V2))+
  geom_point(aes(shape=wdbc_clus$V2,color=wdbc_clus$V2))+
  geom_vline(xintercept=0, linetype="longdash")+
  geom_hline(yintercept=0,linetype="longdash")+
  ggtitle("Unscaled WDBC: Projection of first two eigenvectors (DPP)") + 
  xlab("DPP-PC1") + ylab("DPP-PC2") + labs(color="Groups",shape="Groups") + 
  theme_bw() + theme(panel.border = element_blank())
dev.off()

dpp_iris_test <- sigmat_no_sc(iris)

e <- 130*(pi/180)
test_rotation_iris <- t(matrix(c(cos(e),-sin(e),sin(e),cos(e)),nrow=2))

test_projection_iris<- as.matrix(dpp_iris_test$pro)
test_projection_iris <- t(test_rotation_iris%*%t(test_projection_iris))
test_projection_iris <- as.data.frame(test_projection_iris)

jpeg("test_projection_iris.jpg")
ggplot(test_projection_iris,aes(x=test_projection_iris[,1],y=test_projection_iris[,2],group=iris_clus$V5))+
  geom_point(aes(shape=iris_clus$V5,color=iris_clus$V5))+
  geom_vline(xintercept=0, linetype="longdash")+
  geom_hline(yintercept=0,linetype="longdash")+
  ggtitle("Unscaled Iris: Projection of first two eigenvectors (DPP)") + 
  xlab("DPP-PC1") + ylab("DPP-PC2") + labs(color="Groups",shape="Groups") + 
  theme_bw() + theme(panel.border = element_blank())
dev.off()

dpp_ecoli_test <- sigmat_no_sc(ecoli)

test_projection_ecoli <- as.data.frame(dpp_ecoli_test$pro)

jpeg("test_projection_ecoli.jpg")
ggplot(test_projection_ecoli ,aes(x=test_projection_ecoli [,1],y=test_projection_ecoli [,2],group=ecoli_clus$V9))+
  geom_point(aes(shape=ecoli_clus$V9,color=ecoli_clus$V9))+
  geom_vline(xintercept=0, linetype="longdash")+
  geom_hline(yintercept=0,linetype="longdash")+
  ggtitle("Unscaled E.Coli: Projection of first two eigenvectors (DPP)") + 
  xlab("DPP-PC1") + ylab("DPP-PC2") + labs(color="Groups",shape="Groups") + 
  theme_bw() + theme(panel.border = element_blank())
dev.off()

```

```{r, labels}
library('dplyr')
wdbc_label <- wdbc_clus$V2
wdbc_label[wdbc_label == "M"] <- 1
wdbc_label[wdbc_label == "B"] <- 0
wdbc_label <- as.double(wdbc_label)

iris_label <- data.frame(iris_clus$V5)

iris_label <- iris_label %>%
  mutate(iris_clus.V5 = replace(iris_clus.V5,iris_clus.V5 == "Iris-setosa",0)) %>%
  mutate(iris_clus.V5 = replace(iris_clus.V5,iris_clus.V5 == "Iris-versicolor",1)) %>%
  mutate(iris_clus.V5 = replace(iris_clus.V5,iris_clus.V5 == "Iris-virginica", 2))

ecoli_label <- data.frame(ecoli_clus$V9)

# ecoli_label <- ecoli_label %>%
#   mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "cp",0)) %>%
#   mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "im",1)) %>%
#   mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "imU",2)) %>%
#   mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "om",3)) %>%
#   mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "pp",4))


ecoli_label_test <- ecoli_label %>%
  mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "cp",0)) %>%
  mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "im",1)) %>%
  mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "imU",1)) %>%
  mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "om",1)) %>%
  mutate(ecoli_clus.V9 = replace(ecoli_clus.V9,ecoli_clus.V9 == "pp",1))

```

```{r, ROC curve WDBC}
library('pROC')
#WDBC
#calculate predicted values (Y) with standard PCA, data %*% first PCA eigenvector
Y_wdbc_pca <- as.matrix(wdbc_scaled)%*%as.matrix(pca_wdbc$rotation[,1]) 
wdbc_roc_pca <- roc(wdbc_label,Y_wdbc_pca)

#calculate predicted values with GR-estimated PCA, data %*% first DPP eigenvector
Y_wdbc_dpp <- as.matrix(wdbc_scaled)%*%as.matrix(wdbc_GR$s$v[,1])
wdbc_roc_dpp <- roc(wdbc_label,Y_wdbc_dpp)

jpeg("wdbc_roc.jpg")
plot(x=1-wdbc_roc_pca$specificities,y=wdbc_roc_pca$sensitivities,ylab="True Positives", xlab="False Positives", type="l",col="orange",main="WDBC: ROC Curves")
lines(x=1-wdbc_roc_dpp$specificities,y=wdbc_roc_dpp$sensitivities,type="l",col="blue")
legend(x="bottomright",legend=c("PCA", "DPP"),col=c("orange", "blue"),lty=1:2, cex=0.8)
dev.off()

#calculate AUC
auc_wdbc_pca <- auc(wdbc_label,Y_wdbc_pca)
auc_wdbc_dpp <- auc(wdbc_label,Y_wdbc_dpp)

sprintf("Scaled AUC PCA Method %.4f | Scaled AUC DPP Method %.4f",auc_wdbc_pca,auc_wdbc_dpp)
```

```{r,confusion matrices IRIS}
library(caret)
library(nnet)

#IRIS
#pca
Y_iris_pca <- as.data.frame(as.matrix(iris_scaled)%*%as.matrix(pca_iris$rotation))
Y_iris_pca <- data.frame(Y_iris_pca,iris_clus[5])

model_iris_pca <- multinom(V5~PC1+PC2, data = Y_iris_pca)

p_iris_pca <- predict(model_iris_pca, Y_iris_pca)
tab_iris_pca <- table(p_iris_pca, Y_iris_pca$V5)
tab_iris_pca

iris_pca_mis_error <- 1 - sum(diag(tab_iris_pca)) / sum(tab_iris_pca) # misslassication error / percent
iris_pca_mis_error

#dpp
Y_iris_dpp <- as.matrix(iris_scaled)%*%as.matrix(iris_GR$s$v)
Y_iris_dpp <- data.frame(Y_iris_dpp,iris_clus[5])

model_iris_dpp <- multinom(V5~X1+X2, data = Y_iris_dpp)

p_iris_dpp <- predict(model_iris_dpp, Y_iris_dpp)
tab_iris_dpp <- table(p_iris_dpp, Y_iris_dpp$V5)
tab_iris_dpp

iris_dpp_mis_error <- 1 - sum(diag(tab_iris_dpp)) / sum(tab_iris_dpp) # misslassication error / percent
iris_dpp_mis_error
```

```{r, confusion matrices, WDBC}
#WDBC
#pca
Y_wdbc_pca <- as.data.frame(as.matrix(wdbc_scaled)%*%as.matrix(pca_wdbc$rotation))
Y_wdbc_pca <- data.frame(Y_wdbc_pca,wdbc_clus[2])

model_wdbc_pca <- multinom(V2~PC1+PC2, data = Y_wdbc_pca)

p_wdbc_pca <- predict(model_wdbc_pca, Y_wdbc_pca)
tab_wdbc_pca <- table(p_wdbc_pca, Y_wdbc_pca$V2)
tab_wdbc_pca

wdbc_pca_mis_error <- 1 - sum(diag(tab_wdbc_pca)) / sum(tab_wdbc_pca) # misslassication error / percent
wdbc_pca_mis_error

#dpp
Y_wdbc_dpp <- as.matrix(wdbc_scaled)%*%as.matrix(wdbc_GR$s$v)
Y_wdbc_dpp <- data.frame(Y_wdbc_dpp,wdbc_clus[2])

model_wdbc_dpp <- multinom(V2~X1+X2, data = Y_wdbc_dpp)

p_wdbc_dpp <- predict(model_wdbc_dpp, Y_wdbc_dpp)
tab_wdbc_dpp <- table(p_wdbc_dpp, Y_wdbc_dpp$V2)
tab_wdbc_dpp

wdbc_dpp_mis_error <- 1 - sum(diag(tab_wdbc_dpp)) / sum(tab_wdbc_dpp) # misslassication error / percent
wdbc_dpp_mis_error
```

```{r, confusion matrices, ecoli}
#ECOLI
#pca
Y_ecoli_pca <- as.data.frame(as.matrix(ecoli_scaled)%*%as.matrix(pca_ecoli$rotation))
Y_ecoli_pca <- data.frame(Y_ecoli_pca,ecoli_clus[7])

model_ecoli_pca <- multinom(V9~PC1+PC2, data = Y_ecoli_pca)

p_ecoli_pca <- predict(model_ecoli_pca, Y_ecoli_pca)
tab_ecoli_pca <- table(p_ecoli_pca, Y_ecoli_pca$V9)
tab_ecoli_pca

ecoli_pca_mis_error <- 1 - sum(diag(tab_ecoli_pca)) / sum(tab_ecoli_pca) # misslassication error / percent
ecoli_pca_mis_error

#dpp
Y_ecoli_dpp <- as.matrix(ecoli_scaled)%*%as.matrix(ecoli_GR$s$v)
Y_ecoli_dpp <- data.frame(Y_ecoli_dpp,ecoli_clus[7])

model_ecoli_dpp <- multinom(V9~X1+X2, data = Y_ecoli_dpp)

p_ecoli_dpp <- predict(model_ecoli_dpp, Y_ecoli_dpp)
tab_ecoli_dpp <- table(p_ecoli_dpp, Y_ecoli_dpp$V9)
tab_ecoli_dpp

ecoli_dpp_mis_error <- 1 - sum(diag(tab_ecoli_dpp)) / sum(tab_ecoli_dpp) # misslassication error / percent
ecoli_dpp_mis_error
```

